---
output: html_document
author: Jesse Windle
---

```{r, echo=FALSE}
# To compile this Rmd file, set the working directory in R to the `prelim`
# directory of the repo, then run `rmarkdown::render("prelim_analysis.R")`.
# rmarkdown will compile the file relative to the file's location.
```

```{r, echo=FALSE}
# Packages and external functions
library("ggplot2")
library("gridExtra")
library("readr")
library("dplyr", warn.conflicts = FALSE)
library("lmerTest")
library("kernlab")
source(file.path("..", "shared", "functions.R"))
options(pillar.subtle = FALSE)
```

```{r, echo=FALSE}
depths = -1 * (seq(0, 21, 1) * (16.2/21) + 0.5)
# depths
width = abs(mean(diff(depths)))
depth_ivals = c(depths[1] + 0.5 * width, depths - 0.5 * width)
# depth_ivals
pairwise_depths = 0.5 * (depths[seq(1, 21, 2)] + depths[seq(2, 22, 2)])
# pairwise_depths
```

# RootTracker

RootTracker is a device for non-destructively monitoring root system
architecture.  It has a cylindrical form factor.  Twelve printed
circuit board paddles are arranged in a circular fashion at a fixed
radius and each paddle runs parallel to the z-axis in cylindrical
coordinates.  Running down the paddle are 22 electrodes, which are
used to detect root growth.

RootTracker raw data is collected every 5 minutes.  An algorithm
analyzes that data to identify when a root passes by a sensor.  The
algorithm also keeps track of a device's algorithmic "uptime" which
are the times for which the device is collecting data and that data
passes quality control filters.  Ultimately, data is aggregated to the
hour level, and for a given paddle and electrode RootTracker reports
the number of detections at that electrode and the number of 5 minute
intervals of uptime.

We aggregate the data further in preprocessing so that we record the
number of detections for pairs of electrodes over an entire day.

# Data set

As you can see below, the RootTracker data set we will be working with
has the following attributes:

- `electrode_pair`: the pair of electrodoes running down the paddle, e.g. 1st, 2nd, etc., as an integer
- `paddle`: the paddle, a characeter A-L
- `date`: the date
- `device_id`: the device id
- `count`: the number of detections at that electrode pair, paddle, and day
- `uptime`: the number of 5 minute increments of uptime for that electrode pair, paddle, and day
- `depth_cm`: the average depth of the electrode pair
- `x`: the x location of the device
- `y`: the y location of the device
- `crop`: the crop, e.g. Corn, Wheat, etc.
- `rep`: the block rep, 1 or 2
- `bench`: the bench in the greenhouse
- `days_since_start`: the days since the start of the experiment

The electrodes near the surface can be susceptible to volatile raw
signals.  The same is true for the first few days of an experiment as
soil settles.  Thus, after loading the data, we remove the first pair
of electrodes and the first 4 days.  We also adjust uptime so that it
is measured in days, specifically electrode area-days.

```{r}
df_all = load_rt_data(file.path("..", "data", "acc-1-v3.csv")) %>%
    filter(uptime > 0, electrode_pair > 1, days_since_start > 4) %>%
    mutate(uptime = uptime * 5 / (60 * 24))
```
```{r, echo=FALSE}
knitr::kable(head(df_all, n=5))
```

We further modify the data by assuming that even if there is more than
one detection within a day, for a given paddle and electrode pair
combination, that only corresponds to one root.  Below we will often
aggregate data and then compute a rate of root growth, which is
`count/uptime`.  This lets us normalize the root count by the amount
of algorithmic uptime, effectively accounting for the number of
observations we have over the period of interest, or conversely the
amount of missing data.

```{r}
df_all$old_count = df_all$count
df_all$count = 1 * (df_all$old_count > 0)
df_all$rate = with(df_all, count / uptime)
```


# Root growth over time

We want to understand how root system architecture evolves over time.
We will restrict our attention to how the distribution of root growth
by depth changes over time.  To be clear, we are interested a species'
average root growth, but we expect there to be plant-to-plant
deviations from that average, as well as measurement error, hence
smoothing the data is critical.

## Smoothing by pieces

One approach to smoothing is to aggregate time and depth into
sufficiently large chunks.  Below we group the date into 4 day chunks
and we group depths into the following bins:

```{r, echo=FALSE, comment=""}
depth_bins_coarse = depth_ivals[seq(3, 23, by=4)]
```
```{r}
df_coarse = coarsen_rt_data(
    df_all,
    days_since_start = seq(0, 28, 4),
    depth_cm = depth_bins_coarse
)
```

We have not actually aggregated anything yet, we have just created
larger bins for time and depth, as seen below.

```{r, echo=FALSE}
knitr::kable(head(df_coarse, n=5))
```

We can then aggregate the `count` and `uptime`.  

```{r}
df_time_elec = agg_rt_data(df_coarse, crop, days_since_start, depth_cm)
df_time_elec
```

```{r, echo=FALSE, comment=""}
knitr::kable(df_time_elec[1:5,])
```

We can now plot these averages by depth and time.  For both corn and
wheat, we appear to have shallower root growth, initially, which then
moves deeper later.  We include two dicots here, soybean and cotton,
to point out that they have distinct root growth as compared to the
monocots, with what appears to be a spike of root growth in days 8-12
or 12-16 and then little subsequent activity.  (Below, we will
consider more sophisticated smoothing to produce a higher resolution
picture.)

```{r root-growth-tile, echo=FALSE, fig.width=8, fig.height=4}
time_elec_med = median((df_time_elec %>% filter(crop %in% c("Corn", "Wheat")))$rate)
df_time_elec %>%
    filter(
        crop %in% c("Corn", "Wheat", "Soy", "Cotton"),
    ) %>%
    ggplot() +
    geom_tile(aes(days_since_start, depth_cm, fill=rate)) + 
    scale_fill_gradient2(midpoint=time_elec_med) +
    # scale_fill_gradientn(colours = terrain.colors(10)) +
    facet_wrap(~ crop) +
    theme(axis.text.x = element_text(angle=45, vjust=1.0, hjust=1)) +
    ggtitle("Average rate of root growth by depth, date, and species") +
    guides(fill=guide_legend(title="rate\n(detections/day/\nelectrode area)"))
```

Breaking things out by rep, we see that both reps show similar
patterns of root growth for each species.
```{r}
df_time_elec_rep = agg_rt_data(df_coarse, crop, rep, days_since_start, depth_cm)
df_time_elec_rep %>%
    filter(
        crop %in% c("Corn", "Wheat"),
    ) %>%
    ggplot() +
    geom_tile(aes(days_since_start, depth_cm, fill=rate)) + 
    scale_fill_gradient2(midpoint=time_elec_med) +
    # scale_fill_gradientn(colours = terrain.colors(10)) +
    facet_wrap(rep ~ crop) +
    theme(axis.text.x = element_text(angle=45, vjust=1.0, hjust=1)) +
    ggtitle("Average rate of root growth by depth, date, species, and rep") +
    guides(fill=guide_legend(title="rate\n(detections/day/\nelectrode area)"))

```

As an alternative visualization, we can plot the rate of root growth
across depths, stratified by our time groups.  In both the plots above
and below, we see early root growth at shallower depths, which moves
deeper over time.  The plots below also give us some sense of what the
distribution of root growth by depth should look like.  In the case of
corn, unimodal, with the mode moving deeper over time.

```{r corn-growth-lines, echo=FALSE, fig.width=8, fig.height=4}
df_time_elec %>%
    filter(crop == "Corn") %>%
    ggplot(aes(depth_cm, rate)) +
    geom_path(group=1) +
    facet_wrap(~ days_since_start) +
    theme(axis.text.x = element_text(angle=45, vjust=1.0, hjust=1)) +
    ylab("rate (detect. / day / elect. area)") +
    coord_flip() +
    ggtitle("Rate of root growth by depth, stratified by 'days since start'")
```


### Partitioning variation

As mentioned previously, we are interested in the species-level
average root growth by time and depth, but there are other sources of
variation, like plant-to-plant deviations from that average.  We use a
mixed model to partition this variation, including both time-depth and
time-depth-device random effects.

```{r}
df_coarse_corn = df_coarse %>% filter(crop == "Corn", uptime > 0)
```
```{r}
mm_coarse_corn = lmer(
    rate ~ (1|days_since_start:depth_cm) + (1|days_since_start:depth_cm:device_id),
    data=df_coarse_corn
)
```

As we can see below, the amount of variation in when and where the
roots are growing, for the species average, is of roughly of the same
order, though slightly less than the deviation in that average for
each device.  The residual variance represents not just measurement
error, but differences in when and where detections are occurring
within each bin, so it is not surprising that this number is
relatively large.

```{r, comment="", echo=FALSE}
summary(mm_coarse_corn)
```

Doing the same thing for wheat, we see a similar story.
```{r, echo=FALSE}
df_coarse_wheat = df_coarse %>% filter(crop == "Wheat", uptime > 0)
mm_coarse_wheat = lmer(
    rate ~ (1|days_since_start:depth_cm) + (1|days_since_start:depth_cm:device_id),
    data=df_coarse_wheat
)
```

```{r, comment="", echo=FALSE}
summary(mm_coarse_wheat)
```


## Smoothing via a kernel

We now want to adopt a more sophisticated approach to inferring the
average root growth by time and depth.  In particular, we will employ
a Gaussian process to infer this surface.

For the sake of computational convenience, we still aggregate some,
but not only by two day increments and depths within the following
break points.

```{r, echo=FALSE, comment=""}
depth_bins_coarse_gp = depth_ivals[seq(3, 23, by=2)]
```

```{r}
df_coarse_gp = coarsen_rt_data(
    df_all,
    days_since_start = seq(0, 28, 2),
    depth_cm = depth_bins_coarse_gp
)
```

We then aggregate the count and uptime on that coarser grid and over
all paddles.  We extract coordinates from our depth and time groups
(which are ordered factors).

```{r}
df_dev_time_elec = agg_rt_data(df_coarse_gp, crop, device_id, rep, days_since_start, depth_cm) %>%
    mutate(
        days_coord = as.numeric(days_since_start),
        depth_coord = as.numeric(depth_cm)
    ) %>%
    ungroup()
```

```{r, echo=FALSE}
knitr::kable(head(df_dev_time_elec, n=5))
```

For our Gaussian process model we use a squared exponential kernel
where the distance between points is determined using the date and
depth coordinates defined above.  The length scale is fit to the data
by default for `gausspr`.  Note, we first demean `y`.
```{r}
df_dte_corn = df_dev_time_elec %>% filter(crop == "Corn")
ym = mean(df_dte_corn$rate)
ygp = df_dte_corn$rate - ym
```
```{r}
gp1 = gausspr(x = ~ days_coord + depth_coord, y = ygp, data = df_dte_corn)
```

```{r, echo=FALSE, comment=""}
gp1
```

We can explicitly set the length scale to enforce more smoothing.
```{r}
gp2 = gausspr(x = ~ days_coord + depth_coord, y = ygp, data = df_dte_corn, kpar = list(sigma=0.2))
```

```{r, echo=FALSE}
df_agg_fit = df_dte_corn %>%
    select(days_since_start, depth_cm, days_coord, depth_coord) %>%
    distinct()
df_agg_fit$yhat1 = predict(gp1, df_agg_fit) + ym
df_agg_fit$yhat2 = predict(gp2, df_agg_fit) + ym
```

In either case, a picture of root growth emerges for corn in which
there root growth is shallower initially, and then it moves deeper
over time.

```{r gaussian-process-plots, echo=FALSE, fig.width=12, fig.height=5}
p_gp_1 <- ggplot(df_agg_fit) + 
    geom_tile(aes(days_since_start, depth_cm, fill=yhat1)) +
    scale_fill_gradient2(midpoint=ym) +
    ggtitle("Length scale fit by data") +
    theme(
        plot.title = element_text(size=10),
        axis.text.x = element_text(angle=45, vjust=1.0, hjust=1),
        legend.title = element_text(size=8)
    ) +
    guides(fill=guide_legend(title="rate\n(detect./day/\nelec. area)"))

p_gp_2 <- ggplot(df_agg_fit) + 
    geom_tile(aes(days_since_start, depth_cm, fill=yhat2)) +
    scale_fill_gradient2(midpoint=ym) +
    ggtitle("Length scale chosen for more smoothing") +
    theme(
        plot.title = element_text(size=10),
        axis.text.x = element_text(angle=45, vjust=1.0, hjust=1),
        legend.title = element_text(size=8)
    ) +
    guides(fill=guide_legend(title="rate\n(detect./day/\nelec. area)"))

grid.arrange(p_gp_1, p_gp_2, ncol=2, top="Posterior mean rate of root growth by depth and time for corn")
```
